---
title: "assignment9"
author: "Peter Clark"
date: "March 28, 2017"
output: html_document
---

#DRAFT WORK IN PROGRESS


Read in actual data from Farm Woodlot study at JRF

```{r}
Data<-read.table("farmwoodlot_truncated.csv", header=T, sep=",",stringsAsFactors=F)
str(Data)
```

Create ANOVA function

```{r}
###########################################################################
#Function for a simple ANOVA 
#This fucntion takes continous y and discrete x variables
###########################################################################

myANOVA <- function(x=as.factor(c(rep("maple", 10), rep("oak", 10))), y=runif(20)){
  anovaModel <- aov(y~x)
  
  out <- list(anovaModel, summary(anovaModel))
return(out)
}

###########################################################################
# END OF FUNCITON
###########################################################################
```


Create ANOVA plot function

```{r}

###########################################################################
# Function that plots box plots from ANOVA 
#This fucntion takes continous y and discrete x variables
###########################################################################

myANOVAplot <- function(x=as.factor(c(rep("maple", 10), rep("oak", 10))), y=runif(20)){
  
  aovPlot <- boxplot(y~x, main="ANOVA PLOT")
  
return(aovPlot)
}

###########################################################################
# END OF FUNCITON
###########################################################################
```


Run ANOVA on my data

```{r}
#run my anova on my data
myANOVA(x=Data$PlotNo, y=Data$DBH_48)
```

Plot my data using an ANOVA, comparing DBH by plot/forest type

```{r}
myANOVAplot(x=Data$PlotNo, y=Data$DBH_48)
```


Create simple linear regression function

```{r}
###########################################################################
# Function for a simple linear regression 
# This function takes continous y and x variables
###########################################################################

Regression <- function(x=1:15, y=runif(15)){
  regModel <- lm(y~x)
  
  out <- list(regModel, summary(regModel))
return(out)
}

###########################################################################
# END OF FUNCITON
###########################################################################
```

Create lr plot

```{r}
###########################################################################
# function that plots data from regression 
# this functon takes continous y and x variables
###########################################################################

RegressionPlot <- function(x=1:20, y=runif(20)){
  
  rPlot <- plot(y=y,x=x,pch=21,bg="tan",cex=1.5, main="REGRESSION PLOT")
  regModel <- lm(y~x)
  abline(regModel)

return(rPlot)
}

###########################################################################
# END OF FUNCITON
###########################################################################
```

Clean up data by changing "cull" and blank data to "NA"

```{r}
hist(Data$DBH_63)
#Make all '9999' and '7777' into na
Data$DBH_63[Data$DBH_63 == 9999] <- NA
Data$DBH_63[Data$DBH_63 == 7777] <- NA

hist(Data$DBH_63)

```


Run a regression on tree growth (DBH) over 20 year time period. Yes, simple.

```{r}
Regression(x=Data$DBH_48, y=Data$DBH_63)
```

Plot above regression

```{r}
RegressionPlot(x=Data$DBH_48, y=Data$DBH_63)

```


####Create mock data representative of actual


```{r}
# creating mock data that have the same distribution, mean and stdev of my 
# 1948 dbh data
plotVar <- rep.int((1:4), 5) # four plots
speciesVar <- as.factor(rep(c("TSCA","ACRU","QURU", "FAGR"),each=5)) # four species
library(fGarch) # use this package to get skew normal distribution
dbhVar <- rsnorm(20, mean = 7.1, sd = 2.9, xi = 1.4) # runs the skew normal
# assigns mean, variance and "xi" skew
hist(dbhVar) # random runs. some work some don't
mockDF <- data.frame(plotVar, speciesVar, dbhVar)
print(dataFrame)
```

```{r}
hist(Data$DBH_48)
```

```{r}
dbh48<-na.omit(Data$DBH_48)
mean(dbh48)
sd(dbh48)
is.numeric(Data$DBH_48)
```

