---
title: "Vegan Tutorial"
author: "Peter Clark"
date: "April 12, 2017"
output: html_document
---


##A tutorial in VEGAN
####an R package for community analysis


The **vegan** package provides tools for descriptive community ecology. It has most basic functions of:

* diversity analysis  
* community ordination  
* dissimilarity analysis  

In this tutorial, we will briefly explore the breadth of the program as well as dive into basic diversity analysis explore ordination of multivariate datasets.

If you haven't done so already, please install **vegan**
```{r}
#install.packages("vegan")
```

##Documentation

Consider visiting the [vegan documentation](https://cran.r-project.org/web/packages/vegan/vegan.pdf) to learn about this package.

##Datasets
We will using a few datasets native to vegan

```{r}
library(vegan)
#data(package = "vegan") ## names of data sets in the package

```

##Diversity

Explore the utility of the **diversity** function

```{r}
data(dune) # Vegetation and Environment in Dutch Dune Meadows
str(dune) #a data frame of observations of 30 species at 20 sites

diversity(dune,index = "simpson") # calculate Simpson's 1-D Index of Diversity for each site. # closer to 1 = greater diversity
simpson <- diversity(dune, "simpson") # or assign to var.
simpson 
shannon <- diversity(dune) # note that Shannon's is default
shannon #Typically ranges from 1.5 - 3.4, higher = more diverse 

# lets compare the two
par(mfrow = c(1, 2))  # use par to generate panels with 1 row of 2 graphs
hist(simpson)
hist(shannon)
```

Next we can calcuate all of the pair-wise dissimilarity (distance) measures between sites based on their species composition using the function **vegdist**

Vegdist computes dissimilarity indices. We are using *gower* and *bray-curtis* which are good in detecting underlying ecological gradients

Both indices are used to quantify the compositional dissimilarity between two different sites. They are bounded between 0 and 1, where 0 = same composition, 1 = maximally dissimilar. 

```{r}
# ?vegdist
par(mfrow = c(1, 2))
bray = vegdist(dune, "bray") 
gower = vegdist(dune, "gower")
hist(bray)
hist(gower)

# r allows for multiple iterations of each dissimilarity index to examine #freqeuncy of differences
```

##Rarefaction


Rarefaction is a technique to assess expected species richness. Rarefaction allows the calculation of species richness for a given number of individual samples, based on the construction of rarefaction curves.

The issue that occurs when sampling various species in a community is that the larger the number of individuals sampled, the more species that will be found. Rarefaction curves are created by randomly re-sampling the pool of N samples multiple times and then plotting the average number of species found in each sample (1,2, ... N). "Thus rarefaction generates the expected number of species in a small collection of n individuals (or n samples) drawn at random from the large pool of N samples.". Rarefaction curves generally grow rapidly at first, as the most common species are found, but the curves plateau as only the rarest species remain to be sampled.

To try some rarefaction, we use the **rarefy** and **rarecurve** functions.

```{r}
spAbund <- rowSums(dune)  #gives the number of individuals found in each plot
spAbund # view observations per plot (site 17)
raremin <- min(rowSums(dune))  #rarefaction uses the smallest number of observations per sample to extrapolate the expected number if all other samples only had that number of observations
raremin # view smallest # of obs 

sRare <- rarefy(dune, raremin) # now use function rarefy
sRare #gives an expected number of observations
par(mfrow = c(1, 2))
plot(spAbund, sRare, xlab = "Observed No. of Species", ylab = "Rarefied No. of Species") # plots actual observations by rarefied.
rarecurve(dune, col = "blue") # produces rarefaction curves demonstrating species accumulation 

# notice that rarefied numbers are generally lower than observed, as
# expected notice that they mostly asymptote, indicating thorough sampling
```

##Non-Metric Multidimensional Scaling 
###(aka NMDS, MDS, NMS)


Now let's explore some **ordination techniques**.

Many ordination techniques exist, including principal components analysis (PCA), *non-metric multidimensional scaling (NMDS)*, and correspondence analysis (CA), among others.

Vegan is especially good at NMDS. This tutorial explores this in most detail.

Let's lay some groundwork:

The goal of NMDS is to collapse information from multiple dimensions (e.g, from multiple communities, sites, etc.) into just a few, so that they can be *visualized and interpreted*. 

Consider a single axis representing the abundance of a single species. Along this axis, we can plot the communities in which this species appears, based on its abundance within each.

```{r}

plot(0:10,0:10,type="n",axes=F,xlab="Abundance of Species 1",ylab="")
axis(1)
points(5,0); text(5.5,0.5,labels="community A")
points(3,0); text(3.2,0.5,labels="community B")
points(0,0); text(0.8,0.5,labels="community C")

```

Now consider a second axis of abundance, representing another species. We can now plot each community along the two axes (Species 1 and Species 2).

```{r}
plot(0:10,0:10,type="n",xlab="Abundance of Species 1",
     ylab="Abundance of Species 2")
points(5,5); text(5,4.5,labels="community A")
points(3,3); text(3,3.5,labels="community B")
points(0,5); text(0.8,5.5,labels="community C")
```

Now consider a third axis of abundance representing yet another species.

```{r}
# install.packages("scatterplot3d")
library(scatterplot3d)
d=scatterplot3d(0:10,0:10,0:10,type="n",xlab="Abundance of Species 1",
  ylab="Abundance of Species 2",zlab="Abundance of Species 3"); d
d$points3d(5,5,0); text(d$xyz.convert(5,5,0.5),labels="community A")
d$points3d(3,3,3); text(d$xyz.convert(3,3,3.5),labels="community B")
d$points3d(0,5,5); text(d$xyz.convert(0,5,5.5),labels="community C")
```

The goal of NMDS is to represent the position of communities in multidimensional space as accurately as possible using a reduced number of dimensions that can be easily visualized (and to spare your thinker).

NMDS does not use the absolute abundances of species in communities, but rather their rank orders. The use of ranks omits some of the issues associated with using absolute distance (e.g., sensitivity to transformation), and as a result is much more flexible technique that accepts a variety of types of data. (It's also where the "non-metric" part of the name comes from.)

To run the NMDS, we will use the function **metaMDS**. The function requires only a community-by-species matrix (which we will create randomly).

```{r}
set.seed(2) # random no. generator 
community_matrix=matrix(
   sample(1:100,300,replace=T),nrow=10,
   dimnames=list(paste("community",1:10,sep=""),paste("sp",1:30,sep="")))
head(community_matrix)
example_NMDS=metaMDS(community_matrix, # Our community-by-species matrix
                     k=2) # The number of reduced dimensions. Incress if high stress is problem

```
You should see each iteration of the NMDS until a solution is reached (i.e., stress was minimized after some number of reconfigurations of the points in 2 dimensions).


Now we can plot the NMDS. The plot shows us both the communities ("sites", open circles) and species (red crosses), but we don't know which circle corresponds to which site, and which species corresponds to which cross.

```{r}
plot(example_NMDS)
```

We can use the function ordiplot and orditorp to add text to the plot in place of points to make some sense of this rather non-intuitive mess.

```{r}
ordiplot(example_NMDS,type="n")
orditorp(example_NMDS,display="species",col="red",air=0.01)
orditorp(example_NMDS,display="sites",cex=1.25,air=0.01)

```


Let's suppose that communities 1-5 had some treatment applied, and communities 6-10 a different treatment. We can draw convex hulls connecting the vertices of the points made by these communities on the plot. I find this an intuitive way to understand how communities and species cluster based on treatments.

```{r}
treat=c(rep("Treatment1",5),rep("Treatment2",5))
ordiplot(example_NMDS,type="n")
ordihull(example_NMDS,groups=treat,draw="polygon",col="grey90",label=F)
orditorp(example_NMDS,display="species",col="red",air=0.01)
orditorp(example_NMDS,display="sites",col=c(rep("green",5),rep("blue",5)),
   air=0.01,cex=1.25)
```


One can also plot "spider graphs" using the function orderspider, ellipses using the function ordiellipse, or a minimum spanning tree (MST) using ordicluster which connects similar communities (useful to see if treatments are effective in controlling community structure).
```{r}
#spider plot
ordiplot(example_NMDS,type="n")
ordispider(example_NMDS,groups=treat)
orditorp(example_NMDS,display="species",col="red",air=0.01)
orditorp(example_NMDS,display="sites",col=c(rep("green",5),rep("blue",5)),
         air=0.01,cex=1.25)

#ellipse
ordiplot(example_NMDS,type="n")
ordiellipse(example_NMDS,groups=treat,col="grey",label=F)
orditorp(example_NMDS,display="species",col="red",air=0.01)
orditorp(example_NMDS,display="sites",col=c(rep("green",5),rep("blue",5)),
         air=0.01,cex=1.25)

```

If the treatment is continuous, such as an environmental gradient, then it might be useful to plot contour lines rather than convex hulls. We can simply make up some, say, elevation data for our original community matrix and overlay them onto the NMDS plot using ordisurf:
```{r}
# Define random elevations for previous example
elevation=runif(10,0.5,1.5)
# Use the function ordisurf to plot contour lines
ordisurf(example_NMDS,elevation,main="",col="forestgreen")
# Finally, display species on plot
orditorp(example_NMDS,display="species",col="grey30",air=0.1,
   cex=1)
```

